---
title: "01 · R & RStudio essentials"
subtitle: "Trends in multi-omics data analysis for Microbial Ecology and Biotechnology — UFZ Leipzig"
author: "Instructor: Anderson Santos"
date: "R refresher session"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

## Learning objectives

By the end of this lesson you will be able to:

- Use the **RStudio** panes and run code from a script or an `.Rmd` chunk.
- Create objects with the assignment arrow `<-` and inspect their **type**.
- Work with the workhorse of R — the **vector** — and do maths on whole vectors at once.
- Index and subset vectors, and handle **missing values** (`NA`).
- Write and call a small **function**.
- Chain steps together with the **pipe** (`|>` and `%>%`).

> This is a *refresher*: no prior R needed. Every idea is shown with a runnable
> example you can teach straight from the chunk.

---

## 0 · How R and RStudio fit together

**R** is the language (the engine). **RStudio** is the comfortable dashboard you
drive it from. In RStudio you will mostly use four panes:

- **Source** (top-left): your scripts and `.Rmd` lessons.
- **Console** (bottom-left): where code actually runs; try things here.
- **Environment** (top-right): the objects you have created.
- **Files / Plots / Help** (bottom-right): files, figures, and documentation.

To run one line of a lesson, click inside a grey **code chunk** and press
**Ctrl+Enter** (Windows) / **Cmd+Enter** (Mac). To run a whole chunk, press the
green ▶ button at its top-right.

> 📌 **Teaching note.** Have participants run the very first chunk below live.
> Seeing a number appear in the console is the "hello world" moment that makes R
> feel real.

```{r}
# Anything after a '#' is a COMMENT: R ignores it. Use comments generously.
# The line below is an EXPRESSION. Run it and R prints the result.
2 + 2
```

---

## 1 · Objects and assignment

We store values in **objects** using the assignment arrow `<-`
(read it as "gets"). Typing the object's name prints it.

```{r}
# Store the number of samples in our dataset in an object called n_samples.
n_samples <- 1006          # '<-' means: put 1006 into n_samples

n_samples                  # typing the name prints the value

# You can compute with objects just like with raw numbers.
n_taxa <- 130
n_samples * n_taxa         # total cells in the abundance table
```

R has a few basic **types** of value. The function `class()` tells you which:

```{r}
class(3.14)                # "numeric"  — numbers (with decimals)
class(42L)                 # "integer"  — the L makes it a whole number
class("Firmicutes")        # "character"— text, always in quotes
class(TRUE)                # "logical"  — TRUE / FALSE
```

> 📌 **Teaching note.** Beginners mix up `=` and `<-`. Both can assign, but the
> community convention is `<-` for assignment and `=` only for function
> arguments. Point out RStudio's shortcut **Alt+-** which types ` <- ` for you.

### Exercise 1

Create an object `phylum` holding the text `"Bacteroidetes"`, and an object
`abundance` holding the number `1523`. Then print each one's `class()`.

<details><summary><b>Solution</b> (click to expand)</summary>

```{r}
phylum <- "Bacteroidetes"
abundance <- 1523
class(phylum)      # "character"
class(abundance)   # "numeric"
```
</details>

---

## 2 · Vectors — the heart of R

A **vector** is an ordered collection of values *of the same type*. You build one
with `c()` ("combine"). Almost everything in R is a vector.

```{r}
# Five bacterial genera (a character vector).
genera <- c("Bacteroides", "Prevotella", "Faecalibacterium",
            "Escherichia", "Akkermansia")
genera

# Their (made-up) abundances in one sample (a numeric vector).
counts <- c(1820, 640, 2310, 95, 210)
counts
```

The magic of R: operations apply to **every element at once** (this is called
*vectorisation*) — no loops needed.

```{r}
counts * 2          # double every value
counts / sum(counts)  # convert to proportions (each value / the total)
log(counts)         # natural log of each value
```

Handy summary functions work on the whole vector:

```{r}
length(counts)   # how many elements?
sum(counts)      # total
mean(counts)     # average
max(counts)      # largest
min(counts)      # smallest
```

### 2.1 · Indexing (picking elements)

R counts **from 1** (not 0). Use square brackets `[ ]` to pick elements.

```{r}
genera[1]            # the first genus
genera[c(1, 3)]      # the first and third
counts[counts > 500] # keep only counts above 500 (a logical filter!)
```

That last line is powerful: `counts > 500` produces `TRUE/FALSE`, and R keeps the
`TRUE` positions. You will use this idea constantly.

```{r}
counts > 500                 # the underlying TRUE/FALSE vector
genera[counts > 500]         # which genera are abundant here?
```

### 2.2 · Missing values (`NA`)

Real data has gaps. R marks a missing value as `NA`. Many functions return `NA`
if any input is `NA` — unless you tell them to ignore it with `na.rm = TRUE`.

```{r}
ages <- c(34, 51, NA, 22, 47)   # one unknown age
mean(ages)                      # NA — R refuses to guess
mean(ages, na.rm = TRUE)        # ignore the NA and average the rest
is.na(ages)                     # TRUE exactly where a value is missing
```

> 📌 **Teaching note.** `NA` is *not* zero and *not* the text "NA". It is a
> first-class "unknown". This distinction saves people from silent bugs later.

### Exercise 2

From `counts`, (a) compute the proportion each genus makes of the total, and
(b) find the name of the genus with the **highest** count.
*Hint:* `which.max()` gives the position of the maximum.

<details><summary><b>Solution</b> (click to expand)</summary>

```{r}
counts / sum(counts)          # (a) proportions
genera[which.max(counts)]     # (b) "Faecalibacterium" — the top genus
```
</details>

---

## 3 · Factors — categories with a fixed set of levels

When a character variable represents **categories** (e.g. BMI group), R has a
special type called a **factor**. Factors remember the full set of categories
("levels") and control their order — which matters for tables and plots.

```{r}
bmi <- c("lean", "obese", "lean", "overweight", "obese", "lean")
bmi_f <- factor(bmi, levels = c("lean", "overweight", "obese"))
bmi_f
levels(bmi_f)     # the categories, in the order we chose
table(bmi_f)      # count how many samples fall in each category
```

We will lean on factors heavily when we colour and facet plots later.

---

## 4 · Functions — packaging a recipe

A **function** takes inputs (*arguments*), does something, and returns a result.
You have used built-in ones (`mean`, `sum`). You can write your own with
`function(...) { ... }`. The last line is what it returns.

```{r}
# A function that turns a vector of counts into proportions (%).
to_percent <- function(x) {
  x / sum(x) * 100      # this value is returned
}

to_percent(counts)      # call it just like a built-in function
```

Arguments can have **defaults**:

```{r}
# Shannon diversity index — a real ecology metric we will use all week.
shannon <- function(counts, base = exp(1)) {
  p <- counts[counts > 0] / sum(counts)  # proportions of present taxa
  -sum(p * log(p, base = base))          # the Shannon formula
}

shannon(counts)          # uses the default (natural log)
shannon(counts, base = 2)  # override the default -> "bits"
```

### Exercise 3

Write a function `richness(x)` that returns how many taxa are **present**
(count greater than 0) in a sample. Test it on `c(0, 5, 0, 12, 3)`.

<details><summary><b>Solution</b> (click to expand)</summary>

```{r}
richness <- function(x) {
  sum(x > 0)          # TRUE counts as 1, so this counts the non-zero taxa
}
richness(c(0, 5, 0, 12, 3))   # 3 taxa present
```
</details>

---

## 5 · The pipe — reading code like a sentence

R lets you chain steps with a **pipe**, so `x |> f() |> g()` means "take `x`,
send it to `f`, then send *that* to `g`". It reads left-to-right, like a recipe.

R has two pipes that behave the same for our purposes:

- `|>` — the **native** pipe, built into R (≥ 4.1). Nothing to load.
- `%>%` — the **magrittr** pipe from the tidyverse. You will see it everywhere.

```{r}
# Without a pipe: nested calls, read inside-out (awkward).
round(mean(counts), 1)

# With the native pipe: read left-to-right (natural).
counts |> mean() |> round(1)
```

We use the pipe constantly in the wrangling and plotting lessons, because it
lets us write "filter, then group, then summarise" as one readable flow.

> 📌 **Teaching note.** Type the pipe with **Ctrl+Shift+M**. In RStudio you can
> set *Tools ▸ Global Options ▸ Code* to make that shortcut insert the native
> `|>`.

### Exercise 4

Using the pipe, take `counts`, convert it to proportions with your `to_percent()`
function, and round to 1 decimal — all in one chain.

<details><summary><b>Solution</b> (click to expand)</summary>

```{r}
counts |> to_percent() |> round(1)
```
</details>

---

## Recap

- Store values with `<-`; check their kind with `class()`.
- The **vector** is R's core object; operations are **vectorised** (whole-vector).
- Index with `[ ]` (counting from **1**) and filter with logical conditions.
- Missing data is `NA`; use `na.rm = TRUE` to skip it.
- **Factors** hold ordered categories — key for tables and plots.
- Write reusable **functions**; give arguments **defaults**.
- The **pipe** (`|>` / `%>%`) chains steps into readable, left-to-right flows.

**Next:** in lesson **02** we move from single vectors to whole tables —
importing our real microbiome data as a **tibble** and inspecting it.
