---
title: "01 · R & RStudio essentials"
subtitle: "Trends in multi-omics data analysis for Microbial Ecology and Biotechnology — UFZ Leipzig"
author: "Instructor: Anderson Santos"
date: "R refresher session"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

## Learning objectives

By the end of this lesson you will be able to:

- Use the **RStudio** panes and run code from a script or an `.Rmd` chunk.
- Create objects with the assignment arrow `<-` and inspect their **type**.
- Work with the workhorse of R — the **vector** — and do maths on whole vectors at once.
- Index and subset vectors, and handle **missing values** (`NA`).
- Write and call a small **function**.
- Chain steps together with the **pipe** (`|>` and `%>%`).

> This is a *refresher*: no prior R needed. Every idea is shown with a runnable
> example you can teach straight from the chunk.

---

## 0 · How R and RStudio fit together

**R** is the language (the engine). **RStudio** is the comfortable dashboard you
drive it from. In RStudio you will mostly use four panes:

- **Source** (top-left): your scripts and `.Rmd` lessons.
- **Console** (bottom-left): where code actually runs; try things here.
- **Environment** (top-right): the objects you have created.
- **Files / Plots / Help** (bottom-right): files, figures, and documentation.

To run one line of a lesson, click inside a grey **code chunk** and press
**Ctrl+Enter** (Windows) / **Cmd+Enter** (Mac). To run a whole chunk, press the
green ▶ button at its top-right.

> 📌 **Teaching note.** Have participants run the very first chunk below live.
> Seeing a number appear in the console is the "hello world" moment that makes R
> feel real.

```{r}
# Anything after a '#' is a COMMENT: R ignores it. Use comments generously.
# The line below is an EXPRESSION. Run it and R prints the result.
2 + 2
```

---

## 1 · Objects and assignment

We store values in **objects** using the assignment arrow `<-`
(read it as "gets"). Typing the object's name prints it.

```{r}
# Store the number of samples in our dataset in an object called n_samples.
n_samples <- 1006          # '<-' is R's assignment arrow: put 1006 INTO n_samples
                           # (in R we use '<-', not '='; think of it as "gets")

n_samples                  # type an object's name on its own line -> R prints it

# Objects behave exactly like the values inside them, so we can compute with them.
n_taxa <- 130              # 130 bacterial genera measured in our study
n_samples * n_taxa         # samples x genera = total cells in the abundance table
```

R has a few basic **types** of value. The function `class()` tells you which:

```{r}
# class() reveals the TYPE of a value — getting types wrong is a top beginner bug.
class(3.14)                # "numeric"  -> numbers (decimals); R's default number type
class(42L)                 # "integer"  -> a whole number; the trailing L forces integer
class("Firmicutes")        # "character"-> text; ALWAYS wrapped in quotes, e.g. a taxon name
class(TRUE)                # "logical"  -> TRUE / FALSE, the result of a yes/no test
```

> 📌 **Teaching note.** Beginners mix up `=` and `<-`. Both can assign, but the
> community convention is `<-` for assignment and `=` only for function
> arguments. Point out RStudio's shortcut **Alt+-** which types ` <- ` for you.

### Exercise 1

Create an object `phylum` holding the text `"Bacteroidetes"`, and an object
`abundance` holding the number `1523`. Then print each one's `class()`.

<details><summary><b>Solution</b> (click to expand)</summary>

```{r}
phylum <- "Bacteroidetes"
abundance <- 1523
class(phylum)      # "character"
class(abundance)   # "numeric"
```
</details>

---

## 2 · Vectors — the heart of R

A **vector** is an ordered collection of values *of the same type*. You build one
with `c()` ("combine"). Almost everything in R is a vector.

```{r}
# c() = "combine": glue values into a vector. All elements share ONE type.
# Here: five common human-gut genera as a character (text) vector.
genera <- c("Bacteroides", "Prevotella", "Faecalibacterium",
            "Escherichia", "Akkermansia")
genera                    # print it: R shows the 5 elements in order

# Their (made-up) abundances in one sample — a numeric vector, aligned to 'genera'.
counts <- c(1820, 640, 2310, 95, 210)   # element i is the count of genus i
counts
```

The magic of R: operations apply to **every element at once** (this is called
*vectorisation*) — no loops needed.

```{r}
# Vectorisation: the operation is applied to EVERY element at once, no loop needed.
counts * 2            # multiply each count by 2 -> a new 5-element vector
counts / sum(counts)  # each count / the total = its proportion of the community
log(counts)           # natural log of each value (common for skewed abundance data)
```

Handy summary functions work on the whole vector:

```{r}
# These reductions take the whole vector and return a single summary number.
length(counts)   # how many elements (here: how many genera)?
sum(counts)      # total read count across all genera
mean(counts)     # average count per genus
max(counts)      # the largest count (the most abundant genus)
min(counts)      # the smallest count (the rarest genus)
```

### 2.1 · Indexing (picking elements)

R counts **from 1** (not 0). Use square brackets `[ ]` to pick elements.

```{r}
# Square brackets pick elements by position. R starts counting at 1 (not 0!).
genera[1]            # the 1st genus ("Bacteroides")
genera[c(1, 3)]      # pass a vector of positions -> the 1st and 3rd genera
counts[counts > 500] # logical filter: keep only the counts greater than 500
```

That last line is powerful: `counts > 500` produces `TRUE/FALSE`, and R keeps the
`TRUE` positions. You will use this idea constantly.

```{r}
# The comparison itself produces a TRUE/FALSE vector, one entry per element:
counts > 500                 # TRUE where the count exceeds 500, FALSE otherwise
# Feed that mask to genera[] to keep the names at the TRUE positions:
genera[counts > 500]         # which genera are abundant (>500) in this sample?
```

### 2.2 · Missing values (`NA`)

Real data has gaps. R marks a missing value as `NA`. Many functions return `NA`
if any input is `NA` — unless you tell them to ignore it with `na.rm = TRUE`.

```{r}
ages <- c(34, 51, NA, 22, 47)   # NA marks the one participant whose age is unknown
mean(ages)                      # returns NA — one missing value "poisons" the mean
mean(ages, na.rm = TRUE)        # na.rm = TRUE drops the NA, then averages the rest
is.na(ages)                     # TRUE exactly at the missing position, FALSE elsewhere
```

> 📌 **Teaching note.** `NA` is *not* zero and *not* the text "NA". It is a
> first-class "unknown". This distinction saves people from silent bugs later.

### Exercise 2

From `counts`, (a) compute the proportion each genus makes of the total, and
(b) find the name of the genus with the **highest** count.
*Hint:* `which.max()` gives the position of the maximum.

<details><summary><b>Solution</b> (click to expand)</summary>

```{r}
counts / sum(counts)          # (a) proportions
genera[which.max(counts)]     # (b) "Faecalibacterium" — the top genus
```
</details>

---

## 3 · Factors — categories with a fixed set of levels

When a character variable represents **categories** (e.g. BMI group), R has a
special type called a **factor**. Factors remember the full set of categories
("levels") and control their order — which matters for tables and plots.

```{r}
# Start with plain text categories (one BMI class per participant):
bmi <- c("lean", "obese", "lean", "overweight", "obese", "lean")
# factor() converts text -> a factor; 'levels' fixes the categories AND their order
# (we choose lean < overweight < obese, a meaningful order, not alphabetical).
bmi_f <- factor(bmi, levels = c("lean", "overweight", "obese"))
bmi_f
levels(bmi_f)     # inspect the categories, printed in the order we set
table(bmi_f)      # tally how many participants fall in each category
```

We will lean on factors heavily when we colour and facet plots later.

---

## 4 · Functions — packaging a recipe

A **function** takes inputs (*arguments*), does something, and returns a result.
You have used built-in ones (`mean`, `sum`). You can write your own with
`function(...) { ... }`. The last line is what it returns.

```{r}
# Define a function with function(args){ body }. 'x' is the input placeholder.
# This one turns a vector of counts into percentages of the total.
to_percent <- function(x) {
  x / sum(x) * 100      # last expression = the value the function returns
}

to_percent(counts)      # call it like any built-in; 'counts' becomes x inside
```

Arguments can have **defaults**:

```{r}
# Shannon diversity — a real ecology metric we reuse all week.
# 'base = exp(1)' is a DEFAULT: if the caller omits base, natural log is used.
shannon <- function(counts, base = exp(1)) {
  p <- counts[counts > 0] / sum(counts)  # keep present taxa, convert to proportions
  -sum(p * log(p, base = base))          # the Shannon formula: -sum(p * log p)
}

shannon(counts)            # no 'base' given -> uses the default (natural log, "nats")
shannon(counts, base = 2)  # override the default -> measures diversity in "bits"
```

### Exercise 3

Write a function `richness(x)` that returns how many taxa are **present**
(count greater than 0) in a sample. Test it on `c(0, 5, 0, 12, 3)`.

<details><summary><b>Solution</b> (click to expand)</summary>

```{r}
richness <- function(x) {
  sum(x > 0)          # TRUE counts as 1, so this counts the non-zero taxa
}
richness(c(0, 5, 0, 12, 3))   # 3 taxa present
```
</details>

---

## 5 · The pipe — reading code like a sentence

R lets you chain steps with a **pipe**, so `x |> f() |> g()` means "take `x`,
send it to `f`, then send *that* to `g`". It reads left-to-right, like a recipe.

R has two pipes that behave the same for our purposes:

- `|>` — the **native** pipe, built into R (≥ 4.1). Nothing to load.
- `%>%` — the **magrittr** pipe from the tidyverse. You will see it everywhere.

```{r}
# Without a pipe: calls nest inside each other, so you read INSIDE-OUT (awkward).
round(mean(counts), 1)

# With the native pipe: read LEFT-TO-RIGHT. Read '|>' as the word "then".
counts |> mean() |> round(1)   # take counts, THEN mean(), THEN round to 1 decimal
```

We use the pipe constantly in the wrangling and plotting lessons, because it
lets us write "filter, then group, then summarise" as one readable flow.

> 📌 **Teaching note.** Type the pipe with **Ctrl+Shift+M**. In RStudio you can
> set *Tools ▸ Global Options ▸ Code* to make that shortcut insert the native
> `|>`.

### Exercise 4

Using the pipe, take `counts`, convert it to proportions with your `to_percent()`
function, and round to 1 decimal — all in one chain.

<details><summary><b>Solution</b> (click to expand)</summary>

```{r}
counts |> to_percent() |> round(1)
```
</details>

---

## Recap

- Store values with `<-`; check their kind with `class()`.
- The **vector** is R's core object; operations are **vectorised** (whole-vector).
- Index with `[ ]` (counting from **1**) and filter with logical conditions.
- Missing data is `NA`; use `na.rm = TRUE` to skip it.
- **Factors** hold ordered categories — key for tables and plots.
- Write reusable **functions**; give arguments **defaults**.
- The **pipe** (`|>` / `%>%`) chains steps into readable, left-to-right flows.

**Next:** in lesson **02** we move from single vectors to whole tables —
importing our real microbiome data as a **tibble** and inspecting it.
