---
title: "02 · Data frames & tibbles — importing the microbiome data"
subtitle: "Trends in multi-omics data analysis for Microbial Ecology and Biotechnology — UFZ Leipzig"
author: "Instructor: Anderson Santos"
date: "R refresher session"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

## Learning objectives

- Understand a **data frame / tibble**: a table of columns (variables) and rows (observations).
- **Import** CSV data with `readr::read_csv()` and know why tibbles are friendly.
- **Inspect** a dataset: `glimpse()`, `head()`, `summary()`, dimensions, column types.
- Pull out single columns and turn text columns into **factors**.
- Meet the three tables of our real dataset and how they connect.

---

## 0 · Setup — load packages and find the data

Every lesson starts with this small chunk. It loads the packages we need and
locates the `data/` folder whether you knit from the `notebooks/` folder or from
the project root.

```{r message=FALSE}
library(readr)     # fast CSV import
library(dplyr)     # data wrangling (we use a little here, a lot in lesson 03)
library(tibble)    # modern data frames

# Find the data folder from wherever this file is being run.
data_dir <- if (file.exists("../data/sample_metadata.csv")) "../data" else "data"
```

> 📌 **Teaching note.** If someone gets a "file not found" error, it is almost
> always the working directory. Opening the `.Rproj` fixes this by anchoring the
> project root. The line above makes the lessons robust either way.

---

## 1 · The dataset we will use all session

We use the **HITChip Atlas** — a *real, published* human-gut microbiome study of
**1,006 healthy adults** (Lahti *et al.* 2014, *Nature Communications*). It comes
as three linked tables, the classic shape of omics data:

| Table | File | One row is… | Key columns |
|-------|------|-------------|-------------|
| **Sample metadata** | `sample_metadata.csv` | one participant/sample | `sample_id`, `age`, `sex`, `nationality`, `bmi_group`, `diversity_shannon` |
| **Taxonomy** | `taxonomy.csv` | one bacterial genus | `genus`, `phylum`, `family` |
| **Abundance** | `genus_abundance.csv` | one sample × 130 genera | `sample_id` + 130 abundance columns |

`sample_id` links metadata ↔ abundance; `genus` links taxonomy ↔ abundance
columns. We join them in lesson 03.

---

## 2 · Importing with `read_csv()`

`readr::read_csv()` reads a comma-separated file into a **tibble** (a modern data
frame). It prints a short message telling you the **type it guessed for each
column** — always read that message.

```{r}
meta <- read_csv(file.path(data_dir, "sample_metadata.csv"))
meta
```

Notice what a tibble does for you: it prints only the first 10 rows, shows the
**dimensions** (`1,006 × 7`) and the **type under each column name**
(`<chr>` = text, `<dbl>` = number). That saves you from flooding the console.

> Base R has `read.csv()` too; `read_csv()` (with the underscore) is faster,
> never silently turns text into factors, and gives that helpful type report.

---

## 3 · Inspecting a dataset

Before analysing anything, *look* at it. These are the commands you will run
first on every new table.

```{r}
dim(meta)        # rows, columns
nrow(meta)       # number of rows (samples)
ncol(meta)       # number of columns (variables)
names(meta)      # the column names
```

`glimpse()` is the best single overview: it lists every column, its type, and
the first few values — transposed so wide tables still fit.

```{r}
glimpse(meta)
```

`head()` / `tail()` show the first / last rows; `summary()` gives per-column
statistics (and, crucially, counts of `NA`).

```{r}
head(meta, 4)
summary(meta)
```

> 📌 **Teaching note.** Point at the `summary()` output for `age`: it lists
> `NA's`. Real data is messy — this dataset genuinely has missing ages and sexes,
> which we will handle honestly rather than hide.

### Exercise 1

Import `taxonomy.csv` into an object `taxonomy` and `genus_abundance.csv` into
`abund`. For each, print its **dimensions** and run `glimpse()` (for `abund`,
`glimpse()` will be long — that is fine).

<details><summary><b>Solution</b> (click to expand)</summary>

```{r message=FALSE}
taxonomy <- read_csv(file.path(data_dir, "taxonomy.csv"))
abund    <- read_csv(file.path(data_dir, "genus_abundance.csv"))

dim(taxonomy)   # 130 x 3
dim(abund)      # 1006 x 131 (sample_id + 130 genera)
glimpse(taxonomy)
```
</details>

```{r message=FALSE, include=FALSE}
# (Loaded here so later chunks work even if the exercise was skipped.)
taxonomy <- read_csv(file.path(data_dir, "taxonomy.csv"))
abund    <- read_csv(file.path(data_dir, "genus_abundance.csv"))
```

---

## 4 · Getting at columns

A column is just a vector. Pull one out with `$` (or `[[ ]]`), then use anything
you learned in lesson 01.

```{r}
mean(meta$age, na.rm = TRUE)     # average age (ignoring missing)
range(meta$diversity_shannon)    # min and max diversity
length(unique(meta$nationality)) # how many distinct regions?
```

`table()` counts the categories in a text column — your first look at group sizes.

```{r}
table(meta$nationality)   # samples per geographic region
table(meta$bmi_group)     # samples per BMI class
table(meta$sex, useNA = "ifany")  # include a count of missing sex
```

---

## 5 · Turning text into factors

The category columns (`sex`, `nationality`, `bmi_group`) are text right now. For
grouping, ordered tables and plots, we convert them to **factors** and set a
meaningful order for BMI (lean → morbidly obese, not alphabetical).

```{r}
# Convert with a sensible order for bmi_group.
bmi_levels <- c("underweight", "lean", "overweight",
                "obese", "severeobese", "morbidobese")

meta <- meta |>
  mutate(
    sex         = factor(sex),
    nationality = factor(nationality),
    bmi_group   = factor(bmi_group, levels = bmi_levels)
  )

# Now the levels are ordered the way a clinician would read them.
levels(meta$bmi_group)
table(meta$bmi_group)      # counts respect that order
```

`mutate()` (from **dplyr**) adds or changes columns — we meet it properly next
lesson; here it just lets us overwrite three columns cleanly.

> 📌 **Teaching note.** Ordering a factor once, here, pays off repeatedly: every
> later table and bar chart will list BMI groups in clinical order automatically.

### Exercise 2

(a) How many participants are in each `bmi_group` **after** ordering?
(b) What is the **median** `diversity_shannon` across all samples?
(c) Which `nationality` is the **most common** in the study?

<details><summary><b>Solution</b> (click to expand)</summary>

```{r}
table(meta$bmi_group)                       # (a)
median(meta$diversity_shannon)              # (b)
sort(table(meta$nationality), decreasing = TRUE)[1]  # (c) CentralEurope
```
</details>

---

## 6 · A first peek at the abundance table

The abundance table is **wide**: one row per sample, one column per genus. The
HITChip measures *genus-like groups*, so many column names carry an "et rel."
("and related") suffix, e.g. `Escherichia coli et rel.`. Let's confirm its shape
and look at three of those groups for the first few samples.

```{r}
dim(abund)                              # 1006 samples x 131 columns
abund[1:5, c("sample_id", "Akkermansia", "Escherichia coli et rel.",
             "Bacteroides fragilis et rel.")]
```

Because column names contain characters like spaces, we wrap them in backticks or
quote them when selecting — something we automate with **tidyr** in lesson 03 by
reshaping this wide table into a tidy long one.

---

## Recap

- A **tibble** is a friendly data frame: prints tidily, reports column types.
- Import with `read_csv()`; **read the column-type message** it prints.
- Inspect *before* analysing: `dim()`, `glimpse()`, `head()`, `summary()`.
- A column is a vector — reach it with `$`; count categories with `table()`.
- Convert category text to **factors** and set a meaningful **level order**.
- Our data is three linked tables (metadata, taxonomy, abundance) joined by
  `sample_id` and `genus`.

**Next:** lesson **03** — the wrangling core: `filter`, `select`, `mutate`,
`arrange`, `group_by`/`summarise`, **joins**, and `pivot_longer`/`pivot_wider`.
