---
title: "03 · Data wrangling with dplyr & tidyr"
subtitle: "Trends in multi-omics data analysis for Microbial Ecology and Biotechnology — UFZ Leipzig"
author: "Instructor: Anderson Santos"
date: "R refresher session"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

## Learning objectives

This is the **core** of the session — the toolkit you will reuse all week.

- Reshape the six **dplyr verbs**: `filter`, `select`, `mutate`, `arrange`,
  `summarise`, and `group_by`.
- Combine tables with **joins** (`left_join`).
- Move between **wide** and **long** shapes with `pivot_longer` / `pivot_wider`.
- Tidy text and factors with **stringr** and **forcats** along the way.
- Put it together to compute a real result: **taxonomic composition per region**.

---

## 0 · Setup

```{r message=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(forcats)

data_dir <- if (file.exists("../data/sample_metadata.csv")) "../data" else "data"

meta     <- read_csv(file.path(data_dir, "sample_metadata.csv"))
taxonomy <- read_csv(file.path(data_dir, "taxonomy.csv"))
abund    <- read_csv(file.path(data_dir, "genus_abundance.csv"))

# Order bmi_group once, as in lesson 02.
bmi_levels <- c("underweight","lean","overweight","obese","severeobese","morbidobese")
meta <- meta |> mutate(bmi_group = factor(bmi_group, levels = bmi_levels))

glimpse(meta)
```

---

## 1 · `filter()` — keep the rows you want

`filter()` keeps rows that meet a condition. Conditions use `==`, `>`, `<`, `>=`,
`!=`, and combine with `&` (and), `|` (or).

```{r}
# Participants from Scandinavia.
meta |> filter(nationality == "Scandinavia")

# Lean AND older than 50.
meta |> filter(bmi_group == "lean", age > 50)   # comma = AND

# High-diversity samples (Shannon > 6).
meta |> filter(diversity_shannon > 6) |> nrow()
```

Two very common needs: dropping missing values and matching a **set** of values.

```{r}
meta |> filter(!is.na(age)) |> nrow()              # drop rows with missing age
meta |> filter(nationality %in% c("US", "UKIE"))   # match any in a set
```

### Exercise 1

Keep only participants who are **female**, from **CentralEurope**, and whose
`age` is known. How many are there?

<details><summary><b>Solution</b> (click to expand)</summary>

```{r}
meta |>
  filter(sex == "female", nationality == "CentralEurope", !is.na(age)) |>
  nrow()
```
</details>

---

## 2 · `select()` — keep the columns you want

```{r}
meta |> select(sample_id, age, bmi_group)      # just these columns
meta |> select(-subject_id)                    # everything EXCEPT this one
meta |> select(sample_id, starts_with("div"))  # helper: name starts with "div"
```

`rename()` changes a column name (`new = old`):

```{r}
meta |> rename(shannon = diversity_shannon) |> names()
```

---

## 3 · `mutate()` — add or change columns

`mutate()` creates new columns from existing ones. Every row is computed at once.

```{r}
meta <- meta |>
  mutate(
    # A simple age band using cut(): turns a number into a category.
    age_band = cut(age,
                   breaks = c(0, 30, 45, 60, 100),
                   labels = c("<30", "30-44", "45-59", "60+")),
    # A TRUE/FALSE flag for high diversity.
    high_diversity = diversity_shannon > 5.8
  )

meta |> select(sample_id, age, age_band, diversity_shannon, high_diversity) |> head()
```

`stringr` helps when a column is **text**. Here we make a tidy, lower-case region
label and a short code:

```{r}
meta <- meta |>
  mutate(
    region_label = str_replace_all(nationality, c("UKIE" = "UK & Ireland",
                                                   "US"   = "United States")),
    region_code  = str_sub(nationality, 1, 3) |> str_to_upper()  # first 3 letters
  )

meta |> select(nationality, region_label, region_code) |> distinct() |> head()
```

### Exercise 2

Add a column `diversity_z` giving each sample's Shannon diversity as a
**z-score** (subtract the mean, divide by the standard deviation). Then show the
5 samples with the highest `diversity_z`.
*Hint:* `mean(x)`, `sd(x)`, and `arrange(desc(...))` (next section).

<details><summary><b>Solution</b> (click to expand)</summary>

```{r}
meta |>
  mutate(diversity_z = (diversity_shannon - mean(diversity_shannon)) /
                        sd(diversity_shannon)) |>
  arrange(desc(diversity_z)) |>
  select(sample_id, diversity_shannon, diversity_z) |>
  head(5)
```
</details>

---

## 4 · `arrange()` — sort the rows

```{r}
meta |> arrange(age) |> head(3)            # youngest first
meta |> arrange(desc(diversity_shannon)) |> head(3)  # most diverse first
meta |> arrange(nationality, desc(age)) |> head(3)   # by region, then oldest
```

---

## 5 · `group_by()` + `summarise()` — the heart of analysis

This pattern answers almost every "…per group?" question. `group_by()` splits the
data into groups; `summarise()` collapses each group to one row of statistics.

```{r}
# Average diversity and sample size PER nationality.
meta |>
  group_by(nationality) |>
  summarise(
    n            = n(),                              # count of samples
    mean_shannon = mean(diversity_shannon),          # average diversity
    mean_age     = mean(age, na.rm = TRUE)           # average age (skip NA)
  ) |>
  arrange(desc(mean_shannon))
```

Group by **two** variables for a cross-tabulated summary:

```{r}
meta |>
  filter(!is.na(bmi_group)) |>
  group_by(nationality, bmi_group) |>
  summarise(n = n(), .groups = "drop") |>       # .groups="drop" ungroups after
  head(8)
```

> 📌 **Teaching note.** `summarise()` is where biology appears: "mean diversity
> per region", "richness per BMI class". Emphasise the mental model —
> **split → apply → combine**.

### Exercise 3

For each `bmi_group` (excluding `NA`), report the number of samples and the
**median** Shannon diversity, sorted from most to least diverse.

<details><summary><b>Solution</b> (click to expand)</summary>

```{r}
meta |>
  filter(!is.na(bmi_group)) |>
  group_by(bmi_group) |>
  summarise(n = n(), median_shannon = median(diversity_shannon)) |>
  arrange(desc(median_shannon))
```
</details>

---

## 6 · Reshaping: `pivot_longer()` and `pivot_wider()`

Our abundance table is **wide** (130 genus columns). Most tidyverse tools — and
all of ggplot2 — prefer **long** (tidy) data: one row per *observation*. Here an
observation is "one genus in one sample".

```{r}
abund_long <- abund |>
  pivot_longer(
    cols      = -sample_id,   # everything except the id becomes long
    names_to  = "genus",      # old column NAMES go into this new column
    values_to = "abundance"   # old cell VALUES go into this new column
  )

abund_long          # 1006 samples x 130 genera = 130,780 rows
```

Now every genus/sample pair is a row — perfect for joining, grouping and
plotting. To go back the other way, `pivot_wider()` spreads a long table wide:

```{r}
abund_long |>
  filter(genus %in% c("Akkermansia", "Escherichia coli et rel.")) |>
  pivot_wider(names_from = genus, values_from = abundance) |>
  head()
```

> 📌 **Teaching note.** "Long for computing and plotting, wide for reading and
> matrices." This single idea unlocks most day-to-day omics wrangling.

---

## 7 · Joins — connecting tables

`abund_long` knows the *genus* but not its *phylum*. The `taxonomy` table has
that mapping. `left_join()` glues columns from `taxonomy` onto matching rows of
`abund_long`, matching on the shared `genus` column.

```{r}
abund_tax <- abund_long |>
  left_join(taxonomy, by = "genus")     # add phylum & family columns

abund_tax |> select(sample_id, genus, phylum, abundance) |> head()
```

Now join the **sample metadata** too, matching on `sample_id`, so every row knows
its participant's region and BMI group. This one tidy table combines all three
files:

```{r}
full <- abund_tax |>
  left_join(meta |> select(sample_id, nationality, bmi_group, age),
            by = "sample_id")

glimpse(full)
```

> 📌 **Teaching note.** `left_join` keeps every row of the *left* table and adds
> matching info from the right. If a genus were missing from `taxonomy`, its
> `phylum` would be `NA` — a quick way to catch mismatches.

### Exercise 4

Using `abund_tax`, compute the **total abundance per phylum** across the whole
study, sorted from most to least abundant. Which phylum dominates the human gut
here?
*Hint:* `group_by(phylum)` then `summarise(total = sum(abundance))`.

<details><summary><b>Solution</b> (click to expand)</summary>

```{r}
abund_tax |>
  group_by(phylum) |>
  summarise(total = sum(abundance)) |>
  arrange(desc(total))
# Firmicutes and Bacteroidetes dominate — the classic gut signature.
```
</details>

---

## 8 · Putting it together — composition per region

Now we chain everything into one analysis: **the mean relative abundance of each
phylum, per geographic region.** This is the kind of multi-step pipe you will
write daily.

```{r}
composition <- full |>
  # 1. per sample, total abundance (to make a relative abundance) and per phylum sum
  group_by(sample_id, nationality, phylum) |>
  summarise(phy_abund = sum(abundance), .groups = "drop") |>
  # 2. within each sample, convert to a proportion (relative abundance)
  group_by(sample_id) |>
  mutate(rel_abund = phy_abund / sum(phy_abund)) |>
  ungroup() |>
  # 3. average those proportions within each region
  group_by(nationality, phylum) |>
  summarise(mean_rel = mean(rel_abund), .groups = "drop")

# Show it as a readable wide table (regions as columns), in %.
composition |>
  mutate(mean_pct = round(mean_rel * 100, 1)) |>
  select(phylum, nationality, mean_pct) |>
  pivot_wider(names_from = nationality, values_from = mean_pct)
```

Read across a row to see how a phylum's share shifts between regions — a genuine
ecological pattern, produced with the six verbs plus a join and two pivots. We
**visualise** exactly this table in lesson 04.

```{r include=FALSE}
# Save for reuse / inspection.
saveRDS(composition, file.path(tempdir(), "composition.rds"))
```

### Exercise 5 (challenge)

Reorder the `phylum` factor by its **overall** mean relative abundance (most
abundant first) using `forcats::fct_reorder()`, then show the composition table
again. *Hint:* first compute an overall mean per phylum, or use `fct_reorder`
directly on the long `composition` with `mean_rel`.

<details><summary><b>Solution</b> (click to expand)</summary>

```{r}
composition |>
  mutate(phylum = fct_reorder(phylum, mean_rel, .fun = sum, .desc = TRUE)) |>
  arrange(phylum, nationality) |>
  head(10)
```
</details>

---

## Recap

- **`filter`** rows, **`select`** columns, **`mutate`** new columns,
  **`arrange`** to sort.
- **`group_by` + `summarise`** = split–apply–combine: the answer to every
  "per-group" question.
- **`pivot_longer` / `pivot_wider`** move between tidy-long and readable-wide.
- **`left_join`** connects tables on a shared key (`genus`, `sample_id`).
- **stringr** cleans text, **forcats** orders factors.
- Chaining these with the pipe turns three raw files into a real result:
  **phylum composition per region**.

**Next:** lesson **04** — turn these tables into figures with **ggplot2**.
