---
title: "05 · Capstone — a guided microbiome analysis in R"
subtitle: "Trends in multi-omics data analysis for Microbial Ecology and Biotechnology — UFZ Leipzig"
author: "Instructor: Anderson Santos"
date: "R refresher session"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    df_print: paged
    fig_width: 7
    fig_height: 4.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 110)
```

## The scenario

You have received a human-gut microbiome study — three tables: **abundances**
(1,006 samples × 130 genera), a **taxonomy**, and **participant metadata**. Your
job: take it from raw tables to a few defensible conclusions, combining
everything from lessons 01–04.

## How this notebook works

It is a **hands-on activity**, not a lecture. Each **step** states a task and
gives hints; you write the code in the blank chunk. A **collapsible solution**
sits under each step — try first, peek if stuck. Steps build on each other, so
run them in order.

> ⏱️ Suggested time: 45–60 min. Steps 1–7 are core; 8–10 are extension.
> 📌 **Teaching note.** Run as pair-programming: one drives, one navigates.
> Reveal each solution only after most pairs have attempted the step.

---

## Step 0 · Setup — load packages and data (just run it)

```{r}
library(readr); library(dplyr); library(tidyr)          # import + wrangle
library(ggplot2); library(forcats); library(stringr)    # plot + factors + text
library(vegan)     # community ecology toolkit: diversity() and ordination (metaMDS)

# Locate and load the three tables (same pattern used in every lesson).
data_dir <- if (file.exists("../data/sample_metadata.csv")) "../data" else "data"
meta     <- read_csv(file.path(data_dir, "sample_metadata.csv"))
taxonomy <- read_csv(file.path(data_dir, "taxonomy.csv"))
abund    <- read_csv(file.path(data_dir, "genus_abundance.csv"))

# cat() prints a quick shape report so we can confirm the loads worked.
cat("metadata :", nrow(meta), "x", ncol(meta), "\n")
cat("taxonomy :", nrow(taxonomy), "x", ncol(taxonomy), "\n")
cat("abundance:", nrow(abund), "x", ncol(abund), "\n")
```

---

## Step 1 · Clean the metadata

**Task.** Real data is messy. (a) Report how many rows have a missing `age` and
how many have a missing `sex`. (b) Make an analysis-ready copy `meta_clean` that
keeps only rows with a known `age` **and** known `bmi_group`, and set
`bmi_group` as an **ordered factor** (underweight → morbidly obese).

*Hints:* `is.na()`, `sum()`, `filter()`, `factor(..., levels = ...)`.

```{r step1}
# Your code here

```

<details><summary><b>Solution: Step 1</b></summary>

```{r}
# is.na() flags missing values; sum() counts the TRUEs -> how many gaps in each column.
cat("missing age:", sum(is.na(meta$age)),
    " | missing sex:", sum(is.na(meta$sex)), "\n")

bmi_levels <- c("underweight","lean","overweight","obese","severeobese","morbidobese")
meta_clean <- meta |>
  filter(!is.na(age), !is.na(bmi_group)) |>              # keep only fully-labelled samples
  mutate(bmi_group = factor(bmi_group, levels = bmi_levels))  # + set the clinical order

cat("kept", nrow(meta_clean), "of", nrow(meta), "samples\n")  # show the cost of missing data
```
</details>

```{r echo=FALSE}
# Ensure downstream steps work even if the step was skipped.
bmi_levels <- c("underweight","lean","overweight","obese","severeobese","morbidobese")
meta_clean <- meta |>
  filter(!is.na(age), !is.na(bmi_group)) |>
  mutate(bmi_group = factor(bmi_group, levels = bmi_levels))
```

---

## Step 2 · Reshape abundances to tidy long form

**Task.** Turn the wide `abund` table (one column per genus) into a **long** table
`ab_long` with columns `sample_id`, `genus`, `abundance`. Report its number of
rows and confirm it equals samples × genera.

*Hints:* `pivot_longer(-sample_id, names_to = ..., values_to = ...)`.

```{r step2}
# Your code here

```

<details><summary><b>Solution: Step 2</b></summary>

```{r}
# pivot_longer stacks the 130 genus columns into two: 'genus' (name) + 'abundance' (value).
ab_long <- abund |>
  pivot_longer(-sample_id, names_to = "genus", values_to = "abundance")
# Sanity check: rows should equal samples x genera (ncol - 1 drops the sample_id column).
cat(nrow(ab_long), "rows =", nrow(abund), "samples x", ncol(abund) - 1, "genera\n")
```
</details>

```{r echo=FALSE}
ab_long <- abund |>
  pivot_longer(-sample_id, names_to = "genus", values_to = "abundance")
```

---

## Step 3 · Join taxonomy and metadata

**Task.** Build one tidy table `ab_full` = `ab_long` + phylum/family (from
`taxonomy`, by `genus`) + region/BMI/age (from `meta_clean`, by `sample_id`).
Because `meta_clean` dropped some samples, keep only rows that have metadata
(`inner_join`, or drop `NA` region).

*Hints:* `left_join()` / `inner_join()` on the shared keys.

```{r step3}
# Your code here

```

<details><summary><b>Solution: Step 3</b></summary>

```{r}
ab_full <- ab_long |>
  left_join(taxonomy, by = "genus") |>                    # add phylum/family per genus
  # inner_join keeps ONLY rows that match on both sides -> drops samples cleaned away.
  inner_join(select(meta_clean, sample_id, nationality, bmi_group, age),
             by = "sample_id")
glimpse(ab_full)
```
</details>

```{r echo=FALSE}
ab_full <- ab_long |>
  left_join(taxonomy, by = "genus") |>
  inner_join(select(meta_clean, sample_id, nationality, bmi_group, age),
             by = "sample_id")
```

---

## Step 4 · Alpha diversity per sample (compute it yourself)

**Task.** For every sample compute (a) **observed richness** (number of genera
with abundance > 0) and (b) the **Shannon index** on the abundances. Put them in
`alpha`, then join `meta_clean` and report the **mean of each metric per region**.

*Hints:* group `ab_full` by `sample_id`; richness = `sum(abundance > 0)`;
Shannon = `diversity(abundance, index = "shannon")` from **vegan** (feed it the
per-sample abundance vector), or compute by hand.

```{r step4}
# Your code here

```

<details><summary><b>Solution: Step 4</b></summary>

```{r}
# Group by sample (carrying its metadata) so summarise() gives ONE diversity row per sample.
alpha <- ab_full |>
  group_by(sample_id, nationality, bmi_group, age) |>
  summarise(
    richness = sum(abundance > 0),                       # count genera present (abundance > 0)
    shannon  = diversity(abundance, index = "shannon"),  # vegan's Shannon index on the vector
    .groups  = "drop")

# Then average each metric per region to compare communities.
alpha |>
  group_by(nationality) |>
  summarise(mean_richness = mean(richness),
            mean_shannon  = mean(shannon),
            n = n()) |>                                  # n() = samples per region
  arrange(desc(mean_shannon))                            # most diverse region first
```
</details>

```{r echo=FALSE}
alpha <- ab_full |>
  group_by(sample_id, nationality, bmi_group, age) |>
  summarise(richness = sum(abundance > 0),
            shannon  = diversity(abundance, index = "shannon"),
            .groups  = "drop")
```

---

## Step 5 · A diversity figure

**Task.** Make a **boxplot** of `shannon` across `bmi_group` (from `alpha`),
filled by BMI group, with clear labels and no redundant legend. Bonus: overlay
faint jittered points.

*Hints:* `geom_boxplot()`, `geom_jitter(alpha = ...)`, `labs()`,
`theme(legend.position = "none")`.

```{r step5}
# Your code here

```

<details><summary><b>Solution: Step 5</b></summary>

```{r}
ggplot(alpha, aes(bmi_group, shannon, fill = bmi_group)) +  # category (x) vs number (y)
  geom_boxplot(outlier.alpha = 0.3) +                       # median + quartiles per BMI group
  geom_jitter(width = 0.15, alpha = 0.1, size = 0.4) +      # faint raw points show sample sizes
  labs(title = "Gut diversity across BMI groups",
       x = "BMI group", y = "Shannon diversity") +
  theme_minimal() + theme(legend.position = "none")         # x already labels groups -> no legend
```
</details>

---

## Step 6 · Taxonomic composition per region

**Task.** Compute the **mean relative abundance of each phylum per region** and
draw a **stacked bar chart** (y as %). Name the dominant phylum in each region.

*Hints:* per sample → per-phylum sum → divide by the sample total → average per
region; `geom_col()`, `scale_y_continuous(labels = scales::percent)`.

```{r step6}
# Your code here

```

<details><summary><b>Solution: Step 6</b></summary>

```{r fig.height=5}
# Three-stage split-apply-combine: per-phylum sum -> per-sample share -> region mean.
composition <- ab_full |>
  group_by(sample_id, nationality, phylum) |>
  summarise(phy = sum(abundance), .groups = "drop") |>  # total abundance per phylum in a sample
  group_by(sample_id) |>
  mutate(rel = phy / sum(phy)) |>                       # divide by sample total -> proportion
  group_by(nationality, phylum) |>
  summarise(mean_rel = mean(rel), .groups = "drop")     # average that share per region

ggplot(composition, aes(nationality, mean_rel,
                        fill = fct_reorder(phylum, mean_rel, sum))) +  # order the stack by size
  geom_col() +                                          # stacks automatically (fill varies)
  scale_y_continuous(labels = scales::percent) +        # y axis as %
  labs(title = "Phylum composition across regions",
       x = NULL, y = "Mean relative abundance", fill = "Phylum") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))  # tilt long region labels

# slice_max keeps the single top-ranked phylum within each region group:
composition |> group_by(nationality) |> slice_max(mean_rel, n = 1)
```
</details>

```{r echo=FALSE}
composition <- ab_full |>
  group_by(sample_id, nationality, phylum) |>
  summarise(phy = sum(abundance), .groups = "drop") |>
  group_by(sample_id) |>
  mutate(rel = phy / sum(phy)) |>
  group_by(nationality, phylum) |>
  summarise(mean_rel = mean(rel), .groups = "drop")
```

---

## Step 7 · A simple statistic — does diversity differ by BMI?

**Task.** Test whether Shannon diversity differs across BMI groups. Use a
**one-way ANOVA** (`aov`) of `shannon ~ bmi_group` on `alpha`, and report the
p-value. State in one sentence whether there is evidence of a difference.

*Hints:* `fit <- aov(shannon ~ bmi_group, data = alpha)`; `summary(fit)`.

```{r step7}
# Your code here

```

<details><summary><b>Solution: Step 7</b></summary>

```{r}
fit <- aov(shannon ~ bmi_group, data = alpha)
summary(fit)
# The p-value (Pr(>F)) tells you whether mean diversity differs across BMI
# groups. Interpret it in plain words for the class; pair it with the Step 5 boxplot.
```
</details>

---

## Step 8 · Ordination — NMDS of community structure (extension)

**Task.** Do samples from different regions form different **communities**? Build
a sample × genus matrix, run **NMDS** on Bray–Curtis distances with **vegan**, and
plot the two ordination axes coloured by region.

*Hints:* `pivot_wider()` back to a matrix (drop `sample_id` to a rowname);
`metaMDS(mat, distance = "bray")`; extract `scores(..., display = "sites")`.

```{r step8}
# Your code here

```

<details><summary><b>Solution: Step 8</b></summary>

```{r nmds, results='hide', fig.height=5}
# set.seed() fixes the random draw so everyone in class gets the SAME subset/ordination.
set.seed(1)
# Take a balanced subset: 40 random samples from each of four regions (fast enough live).
sub_ids <- meta_clean |>
  filter(nationality %in% c("Scandinavia","US","SouthEurope","CentralEurope")) |>
  group_by(nationality) |> slice_sample(n = 40) |> pull(sample_id)  # pull() -> a plain vector

# vegan wants a numeric matrix with samples as rows and genera as columns.
mat <- abund |>
  filter(sample_id %in% sub_ids) |>
  tibble::column_to_rownames("sample_id") |>  # move sample_id from a column to row names
  as.matrix()
mat <- mat[rowSums(mat) > 0, ]              # keep only rows (samples) with any reads

# metaMDS: NMDS on Bray-Curtis distances; try = 20 restarts to find a stable solution.
nmds <- metaMDS(mat, distance = "bray", trace = FALSE, try = 20)
```

```{r nmds-plot, fig.height=5}
# scores(..., "sites") = the 2-D coordinates of each sample; turn them into a data frame.
scores_df <- as.data.frame(scores(nmds, display = "sites"))
scores_df$sample_id <- rownames(scores_df)  # recover sample_id from the row names
scores_df <- left_join(scores_df,           # attach each sample's region for colouring
                       select(meta_clean, sample_id, nationality), by = "sample_id")

ggplot(scores_df, aes(NMDS1, NMDS2, colour = nationality)) +  # the two ordination axes
  geom_point(size = 2, alpha = 0.8) +
  stat_ellipse(level = 0.7) +               # a 70% confidence ellipse summarising each region
  labs(title = paste0("NMDS of gut communities (stress = ",
                      round(nmds$stress, 3), ")"),  # stress in the title (lower = better 2-D fit)
       colour = "Region") +
  theme_minimal()
```
</details>

---

## Step 9 · Export your results (extension)

**Task.** Save two outputs: (a) the `alpha` diversity table and (b) the
`composition` table, as CSV files, so a collaborator can use them.

*Hints:* `write_csv(x, "path.csv")`.

```{r step9}
# Your code here

```

<details><summary><b>Solution: Step 9</b></summary>

```{r}
# write_csv() is the inverse of read_csv(): save a tibble back to a shareable CSV file.
write_csv(alpha,       "capstone_alpha_diversity.csv")   # per-sample diversity metrics
write_csv(composition, "capstone_phylum_by_region.csv")  # phylum composition per region
cat("Wrote capstone_alpha_diversity.csv and capstone_phylum_by_region.csv\n")
```
</details>

---

## Step 10 · Conclude

Write **2–3 sentences** answering: *which regions or BMI groups differ most in
their gut communities, and which taxa drive the difference?* Base it on your
figures and the ANOVA. Then note **one thing you would check next** (e.g. adjust
for age, rarefy to equal depth, test region with PERMANOVA `vegan::adonis2`).

> *(Double-click here in RStudio to type your interpretation.)*
>
> - Groups that differ most: …
> - Taxa driving the difference: …
> - One thing I would check next: …

---

## Wrap-up

You just ran a complete microbial-ecology workflow in R: **import → clean →
reshape → join → summarise → visualise → test → ordinate → export**. Every tool
came from lessons 01–04; the only new thing was chaining them into an analysis.

Keep this notebook as a **template**: swap in your own feature table this week and
most of the code still applies.

➡️ Want to see the whole thing written out, with every chunk filled in and all
figures produced? Open **07 · Worked analysis** — the fully worked, read-and-run
version of this capstone.
