---
title: "07 · Worked analysis — a complete microbiome study, start to finish"
subtitle: "Trends in multi-omics data analysis for Microbial Ecology and Biotechnology — UFZ Leipzig"
author: "Instructor: Anderson Santos"
date: "R refresher session"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    df_print: paged
    fig_width: 7
    fig_height: 4.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 110)
```

## What this notebook is

This is a **fully worked, read-and-run analysis** — the *reference* companion to
the fill-in **05 · Capstone**. Every chunk is already written: read it top to
bottom, run it, and watch a complete microbial-ecology study come together on the
real **HITChip Atlas** human-gut dataset. Nothing is hidden behind a blank.

> 📌 **Teaching note.** Use this two ways: (1) as the **instructor's live demo**
> — knit it once, then walk the class through each figure; or (2) as a
> **worked solution** students read *after* attempting the capstone in pairs.
> The capstone (05) asks you to *write* this; here you get to *study* it.

**The arc — the same seven verbs you will repeat all week:**
**import → clean → wrangle → summarise → visualise → test → ordinate → export.**

Everything uses only tools from lessons **01–04** (readr, dplyr, tidyr, ggplot2,
forcats) plus **vegan** for diversity and ordination. CRAN only.

---

## 1 · Import — load the three tables

The study ships as three tidy tables: participant **metadata**, a **taxonomy**
(genus → phylum/family), and a wide **abundance** matrix (1,006 samples × 130
genera).

```{r import}
library(readr); library(dplyr); library(tidyr)
library(ggplot2); library(forcats); library(stringr)
library(vegan)     # community ecology: diversity + ordination

# Works whether you knit from notebooks/ or run from the project root.
data_dir <- if (file.exists("../data/sample_metadata.csv")) "../data" else "data"
meta     <- read_csv(file.path(data_dir, "sample_metadata.csv"))
taxonomy <- read_csv(file.path(data_dir, "taxonomy.csv"))
abund    <- read_csv(file.path(data_dir, "genus_abundance.csv"))

cat("metadata :", nrow(meta),  "x", ncol(meta),  "\n")
cat("taxonomy :", nrow(taxonomy), "x", ncol(taxonomy), "\n")
cat("abundance:", nrow(abund), "x", ncol(abund), "\n")
```

A quick look at what we are working with:

```{r glimpse}
glimpse(meta)
```

---

## 2 · Clean — deal with the missing values

Real data is never complete. Before analysing, we (a) measure the gaps and (b)
build an analysis-ready copy that keeps only samples with a known `age` **and**
`bmi_group`, and turns `bmi_group` into an **ordered factor** so tables and plots
respect the natural underweight → morbidly-obese order.

```{r clean}
# How much is missing?
cat("missing age     :", sum(is.na(meta$age)),       "\n")
cat("missing bmi_group:", sum(is.na(meta$bmi_group)), "\n")
cat("missing sex     :", sum(is.na(meta$sex)),        "\n")

bmi_levels <- c("underweight","lean","overweight",
                "obese","severeobese","morbidobese")

meta_clean <- meta |>
  filter(!is.na(age), !is.na(bmi_group)) |>
  mutate(bmi_group = factor(bmi_group, levels = bmi_levels))

cat("kept", nrow(meta_clean), "of", nrow(meta), "samples after cleaning\n")
```

> 📌 **Teaching note.** Cleaning is not optional busywork — dropping the
> unlabelled rows *now* means every downstream `group_by` is honest. Show the
> class the before/after sample count so they see the cost of missing data.

---

## 3 · Wrangle — reshape and join into one tidy table

The abundance table is **wide** (one column per genus). Most tidyverse verbs want
**long** data, so we `pivot_longer`, then `left_join` the taxonomy (to get each
genus's phylum) and `inner_join` the cleaned metadata (dropping samples we cleaned
away).

```{r wrangle}
ab_long <- abund |>
  pivot_longer(-sample_id, names_to = "genus", values_to = "abundance")

cat(nrow(ab_long), "rows =", nrow(abund), "samples x",
    ncol(abund) - 1, "genera\n")

ab_full <- ab_long |>
  left_join(taxonomy, by = "genus") |>                         # + phylum / family
  inner_join(select(meta_clean, sample_id, nationality, bmi_group, age),
             by = "sample_id")                                 # keep cleaned samples

glimpse(ab_full)
```

---

## 4 · Summarise — alpha diversity per sample

**Alpha diversity** describes how varied a single community is. For every sample
we compute **observed richness** (how many genera are present) and the **Shannon
index** (richness weighted by evenness), then look at the mean per region.

```{r alpha}
alpha <- ab_full |>
  group_by(sample_id, nationality, bmi_group, age) |>
  summarise(
    richness = sum(abundance > 0),                       # genera present
    shannon  = diversity(abundance, index = "shannon"),  # vegan::diversity
    .groups  = "drop")

# Mean diversity per region, most-diverse first.
alpha |>
  group_by(nationality) |>
  summarise(mean_richness = round(mean(richness), 1),
            mean_shannon  = round(mean(shannon), 2),
            n = n()) |>
  arrange(desc(mean_shannon))
```

---

## 5 · Visualise (I) — diversity across BMI groups

The first figure: does gut diversity change across body-mass categories? A
**boxplot** shows the distribution per group; faint jittered points keep us honest
about how many samples sit behind each box.

```{r fig-diversity, fig.height=4.5}
p_div <- ggplot(alpha, aes(bmi_group, shannon, fill = bmi_group)) +
  geom_boxplot(outlier.alpha = 0.3) +
  geom_jitter(width = 0.15, alpha = 0.10, size = 0.4) +
  labs(title = "Gut diversity across BMI groups",
       subtitle = "HITChip Atlas — 1,006 healthy adults",
       x = "BMI group", y = "Shannon diversity") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1))
p_div
```

---

## 6 · Visualise (II) — taxonomic composition per region

The classic microbiome figure. We compute, per sample, each **phylum's** share of
the community, average those shares within each region, and draw a **stacked bar**
(y as %). The Firmicutes/Bacteroidetes balance is the headline.

```{r fig-composition, fig.height=5}
composition <- ab_full |>
  group_by(sample_id, nationality, phylum) |>
  summarise(phy = sum(abundance), .groups = "drop") |>
  group_by(sample_id) |>
  mutate(rel = phy / sum(phy)) |>                     # within-sample proportion
  group_by(nationality, phylum) |>
  summarise(mean_rel = mean(rel), .groups = "drop")   # average per region

p_comp <- ggplot(composition,
                 aes(nationality, mean_rel,
                     fill = fct_reorder(phylum, mean_rel, sum))) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Phylum composition across regions",
       x = NULL, y = "Mean relative abundance", fill = "Phylum") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
p_comp

# The single dominant phylum in each region:
composition |> group_by(nationality) |> slice_max(mean_rel, n = 1)
```

> 📌 **Teaching note.** Point out that **Firmicutes** dominates everywhere —
> exactly what the literature reports for the healthy Western gut — and that the
> Firmicutes/Bacteroidetes ratio is the variable most microbiome papers fixate on.

---

## 7 · Test — is the diversity difference real?

Eyeballing a boxplot is not a statistic. A **one-way ANOVA** asks whether mean
Shannon diversity differs across BMI groups more than we would expect by chance.

```{r anova}
fit <- aov(shannon ~ bmi_group, data = alpha)
summary(fit)
```

Read the **`Pr(>F)`** column: a small p-value (say < 0.05) is evidence that at
least one BMI group differs in mean diversity. Always report the statistic
**together** with the Step 5 boxplot — the picture shows the effect size, the test
shows whether it is distinguishable from noise.

---

## 8 · Ordinate — NMDS of whole communities

Alpha diversity is one number per sample; **beta diversity** compares *whole
communities*. **NMDS** (non-metric multidimensional scaling) on **Bray–Curtis**
distances places each sample in 2-D so that similar communities sit close
together. We colour by region to see whether geography structures the gut
microbiome.

```{r nmds, results='hide', fig.height=5}
# A balanced subset keeps the ordination fast enough for a live class.
set.seed(1)
sub_ids <- meta_clean |>
  filter(nationality %in% c("Scandinavia","US","SouthEurope","CentralEurope")) |>
  group_by(nationality) |>
  slice_sample(n = 40) |>
  pull(sample_id)

mat <- abund |>
  filter(sample_id %in% sub_ids) |>
  tibble::column_to_rownames("sample_id") |>
  as.matrix()
mat <- mat[rowSums(mat) > 0, ]                 # drop any all-zero samples

nmds <- metaMDS(mat, distance = "bray", trace = FALSE, try = 20)
```

```{r fig-nmds, fig.height=5}
scores_df <- as.data.frame(scores(nmds, display = "sites"))
scores_df$sample_id <- rownames(scores_df)
scores_df <- left_join(scores_df,
                       select(meta_clean, sample_id, nationality),
                       by = "sample_id")

p_nmds <- ggplot(scores_df, aes(NMDS1, NMDS2, colour = nationality)) +
  geom_point(size = 2, alpha = 0.85) +
  stat_ellipse(level = 0.7) +                  # 70% confidence ellipse per region
  labs(title = paste0("NMDS of gut communities (stress = ",
                      round(nmds$stress, 3), ")"),
       colour = "Region") +
  theme_minimal(base_size = 12)
p_nmds
```

> 📌 **Teaching note.** A **stress** below ~0.2 means the 2-D picture is a fair
> summary of the true distances. Overlapping ellipses = communities are similar
> across regions; separated ellipses = regional structure. This is the same
> ordination logic as the PCA in lesson 06 — different distance, same idea.

---

## 9 · Export — save the figures and the result tables

A real analysis leaves **artefacts** a collaborator can reuse: the figures and the
summary tables. `ggsave()` writes plots; `write_csv()` writes tables.

```{r export}
# Write figures into the project figures/ folder (created if missing).
fig_dir <- if (dir.exists("../figures")) "../figures" else "figures"
dir.create(fig_dir, showWarnings = FALSE)

ggsave(file.path(fig_dir, "worked_diversity_by_bmi.png"),
       p_div,  width = 7, height = 4.5, dpi = 150)
ggsave(file.path(fig_dir, "worked_phylum_composition.png"),
       p_comp, width = 8, height = 5,   dpi = 150)
ggsave(file.path(fig_dir, "worked_nmds_regions.png"),
       p_nmds, width = 7, height = 5,   dpi = 150)

# Write the two result tables next to this notebook.
write_csv(alpha,       "worked_alpha_diversity.csv")
write_csv(composition, "worked_phylum_by_region.csv")

cat("Saved 3 figures to", normalizePath(fig_dir), "\n")
cat("Wrote worked_alpha_diversity.csv and worked_phylum_by_region.csv\n")
```

---

## Recap — the whole analysis in one breath

We took a real, published human-gut dataset from raw tables to defensible
conclusions:

- **Imported** three tables and **cleaned** away the unlabelled samples.
- **Wrangled** a wide matrix into tidy long form and **joined** taxonomy + metadata.
- **Summarised** alpha diversity, then **visualised** it (boxplot) and the
  **phylum composition** per region (stacked bar).
- **Tested** the diversity difference with an ANOVA and **ordinated** whole
  communities with **NMDS**.
- **Exported** three figures and two result tables.

Every verb came from lessons 01–04; the only new idea was **chaining them into a
study**. Now flip to **05 · Capstone** and write this yourself — or swap in your
own feature table this week, and most of this code still applies.
